---
title: "Step 2- QC filtering and clustering"
---

```{r load libraries}
library(Seurat)
library(ggplot2)
library(gridExtra)
library(stringr)
library(ggplot2)
library(tidyverse) 
 set.seed(7)
```

```{r load and calc % mt}
# object updated date_here
# if starting here, run:
# USER_EDIT 1-- change path below to point to the path to your Seurat object
Seurat_obj<- readRDS(file = "./data/rds/01_rawObject_noFilters_scoresAdded_projectname_date.RDSS")
head(Seurat_obj@meta.data)
```

So, I'm revamping how we look at QC metrics - I find the Violin Plots and FeatureScatters to be too "fuzzy,' so instead let's pull the metadata and look at it through ggplot.
```{r}
### pull data
qc.metrics <- Seurat_obj[[c("nCount_RNA","nFeature_RNA","percent.mt","percent.ribo")]]

### plot w/ ggplot
# % mito
## USER_EDIT 2 -- change the geom_vline 1 and 10 below to be numbers that match where the bulk of your data lies. The black lines in this plot show what will be filtered out and kept in based on your numbers here. Keep track of these numbers because you'll use them in the filtering step 
qc.metrics %>%
  ggplot(aes(percent.mt)) + 
  geom_histogram(binwidth = 0.4, fill="yellow", colour="black") +
  ggtitle("percent.mt") +
  geom_vline(xintercept = c(as.numeric('1'), 10)) ### you can play with these values to get an idea of where the filters should be set...
```

```{r}
# % ribo
## USER_EDIT 3 -- change the geom_vline 13 and 33 below to be numbers that match where the bulk of your data lies. The black lines in this plot show what will be filtered out and kept in based on your numbers here. Keep track of these numbers because you'll use them in the filtering step
qc.metrics %>%
  ggplot(aes(percent.ribo)) + 
  geom_histogram(binwidth = 0.4, fill="yellow", colour="black") +
  ggtitle("percent.ribo") +
  geom_vline(xintercept = c(13, 33))
```

```{r}
# nCount
## USER_EDIT 4 -- change the geom_vline 2500 and 23000 below to be numbers that match where the bulk of your data lies. The black lines in this plot show what will be filtered out and kept in based on your numbers here. Keep track of these numbers because you'll use them in the filtering step
qc.metrics %>%
  ggplot(aes(nCount_RNA)) + 
  geom_histogram(binwidth = 100, fill="yellow", colour="black") +
  ggtitle("nCount_RNA") +
  geom_vline(xintercept = c(2500, 23000))
```

```{r}
# nFeature
## USER_EDIT 5 -- change the geom_vline 2500 and 23000 below to be numbers that match where the bulk of your data lies. The black lines in this plot show what will be filtered out and kept in based on your numbers here. Keep track of these numbers because you'll use them in the filtering step
qc.metrics %>%
  ggplot(aes(nFeature_RNA)) + 
  geom_histogram(binwidth = 100, fill="yellow", colour="black") +
  ggtitle("nFeature_RNA") +
  geom_vline(xintercept = c(1000, 5000))
```
Let's apply some filters to the extracted metadata and check things over...

```{r}
### filter by % mito, % ribo, nFeature, and nCounts 
## store the thresholds as variables for easy use later... 
## USER_EDIT 6-- change the numbers below to match the numbers you put in the graphs above for filtering criterion after visualizing your data 
mito_cutoffs <- c(1,10)
ribo_cutoffs <- c(13, 33)
nFeat_cutoffs <- c(1000, 5000)
nCount_cutoffs <- c(2500, 23000)

qc.metrics_subset <- qc.metrics[qc.metrics$percent.mt > min(mito_cutoffs) & qc.metrics$percent.mt < max(mito_cutoffs) &
                                  qc.metrics$percent.ribo > min(ribo_cutoffs) & qc.metrics$percent.ribo <  max(ribo_cutoffs) &
                                  qc.metrics$nFeature_RNA > min(nFeat_cutoffs) & qc.metrics$nFeature_RNA <  max(nFeat_cutoffs) &
                                  qc.metrics$nCount_RNA > min(nCount_cutoffs) & qc.metrics$nCount_RNA < max(nCount_cutoffs),]

print(paste("# of cells before filters:", nrow(qc.metrics)))
print(paste("# of cells after filters:", nrow(qc.metrics_subset)))
```

```{r}
# % mito
qc.metrics_subset %>%
  ggplot(aes(percent.mt)) + 
  geom_histogram(binwidth = 0.4, fill="blue", colour="black") +
  ggtitle("percent.mt") +
  geom_vline(xintercept = c(min(mito_cutoffs), max(mito_cutoffs))) ### you can play with these values to get an idea of where the filters should be set...

# % ribo
qc.metrics_subset %>%
  ggplot(aes(percent.ribo)) + 
  geom_histogram(binwidth = 0.4, fill="pink", colour="black") +
  ggtitle("percent.ribo") +
  geom_vline(xintercept = c(min(ribo_cutoffs), max(ribo_cutoffs)))

# nCount
qc.metrics_subset %>%
  ggplot(aes(nCount_RNA)) + 
  geom_histogram(binwidth = 100, fill="yellow", colour="black") +
  ggtitle("nCount_RNA") +
  geom_vline(xintercept = c(min(nCount_cutoffs), max(nCount_cutoffs)))

# nFeature
qc.metrics_subset %>%
  ggplot(aes(nFeature_RNA)) + 
  geom_histogram(binwidth = 100, fill="cyan", colour="black") +
  ggtitle("nFeature_RNA") +
  geom_vline(xintercept = c(min(nFeat_cutoffs), max(nFeat_cutoffs)))
```

USER_EDIT 7-- 
Go back up to the filtering of the dataframe and tweak those values until you're happy with these histograms!

Once that's all set, check the Violin and Feature Scatter plots before/after filtering the object based on these ggplot criteria...

Remember, we set some variables above for the filters - let's use those in the Seurat subset call

```{r apply filters and plot QC metrics}
### does the subset approach look familiar?
cached_data  <- subset(Seurat_obj, 
                       subset = percent.mt > min(mito_cutoffs) & percent.mt < max(mito_cutoffs) &
                                  percent.ribo > min(ribo_cutoffs) & percent.ribo <  max(ribo_cutoffs) &
                                  nFeature_RNA > min(nFeat_cutoffs) & nFeature_RNA <  max(nFeat_cutoffs) &
                                  nCount_RNA > min(nCount_cutoffs) & nCount_RNA < max(nCount_cutoffs))
```

```{r}
Idents(Seurat_obj) <- "gem.group"
### violinplots
VlnPlot(Seurat_obj, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2, pt.size=0.1)
VlnPlot(Seurat_obj, features = c("percent.mt", "percent.ribo"), ncol = 2, pt.size=0.1)
### featurescatters
FeatureScatter(Seurat_obj, feature1 = "nCount_RNA", feature2 = "percent.mt", pt.size=0.1, group.by = "gem.group")
FeatureScatter(Seurat_obj, feature1 = "nCount_RNA", feature2 = "percent.ribo", pt.size=0.1, group.by = "gem.group")
FeatureScatter(Seurat_obj, feature1 = "nFeature_RNA", feature2 = "percent.mt", pt.size=0.1, group.by = "gem.group")
FeatureScatter(Seurat_obj, feature1 = "nFeature_RNA", feature2 = "percent.ribo", pt.size=0.1, group.by = "gem.group")
FeatureScatter(Seurat_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", pt.size=0.1, group.by = "gem.group")
```

```{r}
# plot filtered data
# violinplots
VlnPlot(cached_data, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2, pt.size=0.1)
VlnPlot(cached_data, features = c("percent.mt", "percent.ribo"), ncol = 2, pt.size=0.1)
### featurescatters
FeatureScatter(cached_data, feature1 = "nCount_RNA", feature2 = "percent.mt", pt.size=0.1, group.by = "gem.group")
FeatureScatter(cached_data, feature1 = "nCount_RNA", feature2 = "percent.ribo", pt.size=0.1, group.by = "gem.group")
FeatureScatter(cached_data, feature1 = "nFeature_RNA", feature2 = "percent.mt", pt.size=0.1, group.by = "gem.group")
FeatureScatter(cached_data, feature1 = "nFeature_RNA", feature2 = "percent.ribo", pt.size=0.1, group.by = "gem.group")
FeatureScatter(cached_data, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", pt.size=0.1, group.by = "gem.group")
```


```{r}
Seurat_obj <- cached_data
rm(cached_data)
```
##### SCTransform
```{r}
gdata::keep(Seurat_obj, sure = TRUE)
head(Seurat_obj@meta.data)
```

```{r}
### SCTransform 
Seurat_obj <- SCTransform(Seurat_obj, assay = "RNA", new.assay.name = "SCT", 
                                    variable.features.n = 4000, 
                                    variable.features.rv.th = 1.3, 
                                    vars.to.regress = c("percent.ribo", "percent.mt", "G2M.Score", "S.Score"), 
                                    return.only.var.genes = TRUE,conserve.memory= TRUE)

### PCA
Seurat_obj <- RunPCA(Seurat_obj, verbose = FALSE, npcs = 100)
### harmony
Seurat_obj <- harmony::RunHarmony(Seurat_obj, group.by.vars = "gem.group", assay.use="SCT")
### elbowplot, just to check
ElbowPlot(Seurat_obj, ndims = 50)
```

```{r}
### set dims
dims <- 1:35 ### relatively high # of dims, UMAP can take it.
### NOTE: n.components = 3l is for 3d plotting *cool_emoji*)
### umap, neighbors, and clusters
Seurat_obj <- RunUMAP(Seurat_obj, reduction = "harmony", dims = dims) # , n.components = 3L)
Seurat_obj <- FindNeighbors(Seurat_obj, reduction = "harmony", dims = dims)
Seurat_obj <- FindClusters(Seurat_obj, resolution = 0.25)

DimPlot(Seurat_obj, split.by = "gem.group")
DimPlot(Seurat_obj)
```


```{r}
## USER_EDIT 8-- change path to the path name / location you wish to use for this object, an example format is below 
saveRDS(Seurat_obj, "../data/rds/02_SCTandHarmonyObject_FilteredAndClustered_onlyVarGenes_projectname_date.RDS")
```

```{r}
# uncomment this below if you want to start from this step 
#Seurat_obj <- readRDS("../data/rds/02_SCTandHarmonyObject_FilteredAndClustered_onlyVarGenes_projectname_date.RDS")
DimPlot(Seurat_obj, label = TRUE)

VlnPlot(Seurat_obj, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2, pt.size=0.1)
VlnPlot(Seurat_obj, features = c("percent.mt", "percent.ribo"), ncol = 2, pt.size=0.1)
```

Ok, now let's drop low quality clusters (if any) and finish this initial processing.
```{r}
VlnPlot(Seurat_obj, features = "nFeature_RNA")
VlnPlot(Seurat_obj, features = "nCount_RNA")
# uncomment these ggsave lines if you wish to save your plots as images 
# ggsave("../results/01_initial_clustering/01_FirstClustering_nFeature_date.png", width = 12, height = 24, device = "png")
DimPlot(Seurat_obj, label = TRUE)
# ggsave("../results/01_initial_clustering/01_FirstClustering_Dimplot_date.png,", width = 12, height = 12, device = "png")
```
USER_EDIT 9-- based on the plots above, choose which clusters to keep and which to drop. Edit the lines below accordingly-- an example below is shown where there are 12 clusters and clusters 9, 10,and 11 are dropped, but change this for your data! 
```{r}
clusters2keep <- 0:11
clusters2keep <- clusters2keep[clusters2keep != 9 &
                                 clusters2keep != 10 &
                                 clusters2keep != 11]
Seurat_obj <- subset(Seurat_obj, idents = clusters2keep)
```

```{r}
### SCTransform - same settings as before
Seurat_obj <- SCTransform(Seurat_obj, assay = "RNA", new.assay.name = "SCT", 
                                    variable.features.n = 4000, 
                                    variable.features.rv.th = 1.3, 
                                    vars.to.regress = c("percent.ribo", "percent.mt", "G2M.Score", "S.Score"), 
                                    return.only.var.genes = TRUE,conserve.memory=TRUE)

### PCA
Seurat_obj <- RunPCA(Seurat_obj, verbose = FALSE, npcs = 100)
### harmony
Seurat_obj <- harmony::RunHarmony(Seurat_obj, group.by.vars = "gem.group", assay.use="SCT")
### elbowplot, just to check
 ElbowPlot(Seurat_obj, ndims = 50)

### set dims
dims <- 1:35 ### relatively high # of dims, UMAP can take it.
### NOTE: n.components = 3l is for 3d plotting *cool_emoji*)
### umap, neighbors, and clusters
Seurat_obj <- RunUMAP(Seurat_obj, reduction = "harmony", dims = dims) # , n.components = 3L)
Seurat_obj <- FindNeighbors(Seurat_obj, reduction = "harmony", dims = dims)
Seurat_obj <- FindClusters(Seurat_obj, resolution = 0.25)

DimPlot(Seurat_obj, split.by = "gem.group")
DimPlot(Seurat_obj)

VlnPlot(Seurat_obj, features = "nFeature_RNA")
VlnPlot(Seurat_obj, features = "nCount_RNA")
# uncomment these ggsave lines if you wish to save your plots as images and change the path name to match date
#ggsave("../results/01_initial_clustering/02_SecondClustering_dropLowQual_nFeature_date.png", width = 12, height = 24, device = "png")
DimPlot(Seurat_obj, label = TRUE)
# uncomment these ggsave lines if you wish to save your plots as images and change the path name to match date
#ggsave("../results/01_initial_clustering/02_SecondClustering_dropLowQual_Dimplot_date.png,", width = 12, height = 12, device = "png")
```


```{r}
### save RDS!
## USER_EDIT 10-- change path name to the name and location you want to store this object, an example format is below 
saveRDS(Seurat_obj, file = "../data/rds/03_SCTandHarmonyObject_FilteredAndReclustered_onlyVarGenes_orojectname_date.RDS")
### find cluster markers
## parallelize
## USER_EDIT 11-- change to match your compute environment! If you have a mac, you can find this information by clicking the apple in the far top left corner, clicking "About This Mac", system report (button down page under text). The relevant info is  Total Number of Cores:	x. The RAM size can be found in "About This Mac" under Memory (32 GB, for example). If you have 12 cores, make the number of workers = 6 for parallelization. If you have 8, make the number of workers 4, etc. 
future::plan("multisession", workers = 6) ### 12-core // 128gb RAM
options(future.globals.maxSize = 3000 * 1024^2) ### global var size increase - 3gb limit
future::plan()

FoundMarkers <- FindAllMarkers(Seurat_obj, random.seed = 7)
future::plan("sequential") ### 12-core // 128gb RAM

## USER_EDIT 12-- change to path name you wish to home your FoundMarkers csv file, an example format is below 
write.csv(FoundMarkers, "../results/01_initial_clustering/01_FindAllMarkers_harmony_projectname_date.csv")

```

